Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex
Minor edits to algorithm

$\alpha = 1.0$ means that the optimal query should have the fastest runtime (at the expense of quality), while $\alpha=0.0$ means that the optimal query should lose
as little information as possible (at the expense of performance).

@@

Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties, and $N$ counts the number of imputation operations.
For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
the product of the number of attributes and the number of tuples. Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. We wanted to encode the fact that the number of training observations has diminishing returns for imputation quality,
and found that scaling the cell count by applying square root worked well in practice.

To add an alternate imputation strategy to \ProjectName{}, the analyst must provide a corresponding loss with a reasonable parameterization.

@@

We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the I/O cost
per page. The time computation for \textit{Impute} depends on the properties of the underlying
algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
and simply queries it for a time estimate. In the base implementation of our system,
we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
is a function of the number of attributes with no missing data, number of attributes with
missing data being imputed, number of tuples, and the number of iterations for the algorithm.

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire set of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example,~\cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). The \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

@@

\subsubsection{Plan cache}
Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dirty-set specific dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
allows us to obtain the final set of plans that best trade computation
cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

@@

\ProjectName{} is designed so that any imputation strategy, typically a predictive machine
learning model,
can be plugged in with minimal effort and without changes to the optimizer, simply by adding
time and loss functions (\Cref{sec:cost-model}) which describe the new algorithm.
An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing
attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.

As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
based on chained-equation decision trees.  Chained-equation
imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
a predictive model of one missing attribute, conditional on all complete attributes
and all other missing attributes. Chained-equation decision trees algorithms are found to be
effective~\cite{akande2015empirical} and are widely used in epidemiological
domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
implementation~\cite{witten2016data}.

The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
decision trees to a subset of the data and the target imputation attribute. In each
iteration, the missing values of a single attribute are replaced with newly imputed values.
In one {\it epoch} of the algorithm, each of the target imputation
attributes is filled in with imputed values in a single iteration.
With each {\it epoch}, the quality of the imputation improves as values progressively
reflect more accurate relationships between attributes. The algorithm terminates when
convergence is achieved (i.e.\ the imputed values do not change across epochs) or a fixed
number of epochs is reached.

These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset~
\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though in
further work, multiple imputation could be another avenue of assessing
imputation confidence.

>>>>

If $\alpha=0.0$, then the optimal query should lose as little information as possible (at
the expense of performance). If $\alpha = 1.0$, then the optimal query should have the
fastest runtime (at the expense of quality).

@@

Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties, and $N$ counts the number of imputation operations.
For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
the product of the number of attributes and the number of tuples. Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. We encode the fact that the number of training observations has diminishing returns for imputation quality,
and find that scaling the cell count by applying square root works well in practice.

@@

We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the I/O cost
per page. The time computation for \textit{Impute} depends on the properties of the underlying
algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
and simply queries it for a time estimate. In our reference implementation,
we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
is a function of the number of attributes with no missing data, number of attributes with
missing data being imputed, number of tuples, and the number of iterations for the algorithm.

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire set of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which do
not consider CPU cost explicitly). For example,~\cite{martin1995time} finds that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). The \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree.
Nonetheless, in practice, we find that a simple parameterization can be acceptable (i.e.
yield intelligent query plans).

@@

\subsubsection{Plan cache}
Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dirty set-specific dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
allows us to obtain the final set of plans that best trade-off computation
cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

@@

\ProjectName{} is designed so that any imputation strategy
can be plugged in with minimal effort and without changes to the optimizer, simply by adding
time and loss functions (\Cref{sec:cost-model}) which describe the new algorithm.
For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing
attributes $C$ in a relation $R$, returning a new set of tuples $R'$.
The flexibility of this approach aligns with the spirit of query optimization, which aims to
reconcile performance with a declarative interface to data manipulation.

In principle, new imputation algorithms can be added, possibly targeted to a specific
domain. However, \ProjectName{}'s planner will consider an arbitrary subset of the
attributes in any imputation, thus it is feasible to customize the \textit{type} of
algorithm but not necessarily a specific modelling relationships between attributes.

As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
based on chained-equation decision trees.  Chained-equation
imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes being imputed by iteratively fitting
a predictive model of one missing attribute, conditional on all complete attributes
and all other missing attributes. Chained-equation decision trees algorithms are found to be
effective~\cite{akande2015empirical} and are widely used in epidemiological
domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
implementation~\cite{witten2016data}.

The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
decision trees for the target imputation attribute on a subset of the data. In each
iteration, the missing values of a separate attribute are replaced with newly imputed values.
In one {\it epoch} of the algorithm, each of the target imputation
attributes is filled in with imputed values in a single iteration.
With each epoch, the quality of the imputation improves as values progressively
reflect more accurate relationships between attributes. The algorithm terminates when
convergence is achieved (i.e.\ the imputed values do not change across epochs) or a fixed
number of epochs is reached.

These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset~
\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
multiple imputation could incorporated as another avenue of assessing imputation confidence.
